---
title: "Impact of Severe Weather Events on Public Health and Economy in the United States"
subtitle: "Reproducible Research"
Author: "Erik Cornelsen"
output: 
  html_document:
    keep_md: true
    fig_caption: yes
    number_sections: yes
    toc: yes
---

# Synopsis

Storms and other severe weather events can cause both public health and economic problems for communities and municipalities. Many severe events can result in fatalities, injuries, and property damage, and preventing such outcomes to the extent possible is a key concern.  This report is meant to provide guidance on where to prioritize resources to prepare for and deal with different types of severe weather events.

Data comes from the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database. This database tracks characteristics of major storms and weather events in the United States, including when and where they occur, as well as estimates of any fatalities, injuries, and property damage.

During this analysis we looked at differnt types of storm events (as indicated in the `EVTYPE` variable) and found that: 

1. Across the United States, the types of events most harmful with respect to **population health** are TORNADO, EXCESSIVE HEAT, FLOOD, FLASH FLOOD, and LIGHTNING.
1. Across the United States, the types of events which have the greatest **economic consequences** are FLOOD, TORNADO, HAIL, FLASH FLOOD, and DROUGHT.


# Data

## Data Overview

The data for this analysis was downloaded from this [Storm Data Link](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2) [47Mb]. The format is a comma-separated-value file compressed via the bzip2 algorithm to reduce its size.  Below are some links to the available documentation on how some of the variables are constructed/defined:

* National Weather Service [Storm Data Documentation](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf)
* National Climatic Data Center Storm Events [FAQ](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2FNCDC%20Storm%20Events-FAQ%20Page.pdf)
* [Storm Events NICAR Website](http://ire.org/nicar/database-library/databases/storm-events/)
* [Storm Events Data Layout Document](http://ire.org/media/uploads/files/datalibrary/samplefiles/Storm%20Events/layout08.doc)

## Data Processing
Following are the steps taken to get the data in a usable format to try to answer the desired questions.  

Load needed libraries.
```{r init, collapse=TRUE, message=FALSE}
# set global option to echo everything so that code is visible.
library(knitr)
knitr::opts_chunk$set(echo = TRUE, collapse = TRUE, cache=TRUE, cache.path="cache/", fig.width=10, fig.path="figure/")
library(R.utils)
library(data.table)
library(lubridate)
library(dplyr)
library(ggplot2)
library(reshape2)
#library(utils)
library(scales)
sessionInfo()

```

Download and Unzip the Storm Data.
```{r getData, collapse=TRUE, eval=TRUE}
zipFileName <- "data/redata-pa2-StormData.csv.bz2"
unzFileName <- "data/redata-pa2-StormData.csv"
print("downloading  File  ...")
download.file("http://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2", destfile = zipFileName)
#print("unzipping  File  ...")
#bunzip2(filename = zipFileName, destname = unzFileName, overwrite=TRUE, remove=FALSE)
```

```{r unzipData}
unzFileName <- "data/redata-pa2-StormData.csv"
tstFileName <-"data/newlineIssueTest.csv"
stormData <- fread(input = tstFileName,  sep = ",", header=TRUE,
                   na.strings = c("?","NA","N/A","<NA>",""),
                   nrows = -1 )
tail(stormData,3)
sessionInfo()
```


Load the Storm data.   
```{r loadData, collapse=TRUE, cache=TRUE}
#This would be faster using fread but the file contains a couple of line breaks in quoted fields that it can't handle. 
#stormData <- fread(input = unzFileName,  sep = ",", header=TRUE,
#                   select=c("EVTYPE", "BGN_DATE", "FATALITIES", "INJURIES", "PROPDMG", "PROPDMGEXP", "CROPDMG", "CROPDMGEXP"), 
#                   na.strings = c("?","NA","N/A","<NA>",""),
#                   nrows = -1 )

stormData <- read.csv(bzfile(zipFileName), 
                      header = TRUE, 
                      stringsAsFactors = FALSE, 
                      na.strings = c("?","NA","N/A","<NA>",""),  
                      nrows = -1)  #to read in ALL rows (-1)
#str(stormData)
```

The events in the data start in the year 1950 and end in November 2011. Looking at the histogram below we can see that in the earlier years of the database there are generally fewer events recorded, most likely due to a lack of good records. More recent years should be considered more complete.  The number of events tracked starts to significantly increase around 1995. 

```{r histogram}
stormData$YEAR <- as.numeric(format(as.Date(stormData$BGN_DATE, format = '%m/%d/%Y'), "%Y"))
hist(stormData$YEAR, breaks = 30)

#limit data to desired years and desired variables
cntFull <- nrow(stormData)
stormDataReduced <- stormData[stormData$YEAR >= 1995, c("EVTYPE", "BGN_DATE", "YEAR", "FATALITIES", "INJURIES", "PROPDMG", "PROPDMGEXP", "CROPDMG", "CROPDMGEXP")]
cntReduced <- nrow(stormDataReduced) 

```

So, we'll use the subset of the data from 1995 to 2011 to get most out of good records which reduces our observations from `r cntFull` to `r cntReduced`.  

Also discovered that the EVTYPE field (which we want to use to classify the different storms) is **extremly** messy.  When using it as a factor we find 900+ levels.  We'll do a partial cleanup here to reduce the noise but a more thourough treatment is needed to get the best results.
```{r EVTYPE}
# look at EVTYPES and find out why there's 900+ levels
#levels(stormDataReduced$EVTYPE)

# Cleanup: trim whitespace, to upper case, remove any trailing 'S' on EVTYPE to reduce number of factors
cntReduced1 <- nrow(stormDataReduced)
stormDataReduced$EVTYPE <- as.factor(gsub("S$|-$", '', toupper(trim(stormDataReduced$EVTYPE))))

# identify with EVTypes account for the majority of the data (95-98%)
evtypePercentages <- cumsum(sort(table(stormDataReduced$EVTYPE), decreasing = TRUE)) / nrow(stormDataReduced)

# reduce the data to the top 30 EVTYPES, this accounts fr 97.8% of the data
#stormDataReduced <- stormDataReduced[stormDataReduced$EVTYPE %in% names(evtypePercentages[evtypePercentages<.95]),]
stormDataReduced <- stormDataReduced[stormDataReduced$EVTYPE %in% names(evtypePercentages[1:30]),]
cntReduced2 <- nrow(stormDataReduced) 

```

We found that 97.8% of all the data is accounted for with 30 0f the 900+ EVTYPES. So after using just those 30 and filtering out the rest we reduce our rowcount from `r cntReduced1` to `r cntReduced2`.  

Next we need to calculate some additional variables so that we will be able to look at the question of economic impact.  To do so I created a helper function to look at the multiplier variable and calculate the actual value of the cash impact.

* PROPDMG       -	Property damage in whole numbers and hundredths
* PROPDMGEXP	-	A multiplier where Hundred (H), Thousand (K), Million (M), Billion (B)
* PROPCASH	    -	Combines the PROPDMG and PROPDMGEXP fields to create a numeric value (We create this field)
* CROPDMG	    -	Crop damage in whole numbers and hundredths
* CROPDMGEXP	-	A multiplier where Hundred (H), Thousand (K), Million (M), Billion (B)
* CROPCASH	    -	Combines the CROPDMG and CROPDMGEXP fields to create a numeric value (We create this field)
* TOTCASH		-	Combines the PROPCASH and CROPCASH fields to create a numeric value (We create this field)

```{r calcCash}

calcCash <- function(val,exp) {
    if (!is.numeric(val)) stop("val must be numeric")
    
    #used check below to catch a lot of interesting bad data (+, numbers, NA, etc). but then I implemented a switch statement and the default case takes care of all those scenarios by setting cash to $0
    #no longer need this validation.
    #if (!toupper(exp) %in% c("H","K","M","B",NA)) stop(paste0("Invalid 'exp' value: [",exp,"]"))
    
    # h -> hundred, k -> thousand, m -> million, b -> billion
    switch(toupper(exp), 
            H = val * 100, 
            K = val * 1000, 
            M = val * 1000000, 
            B = val * 1000000000, 
            0)
}

stormDataReduced$PROPCASH <- mapply(calcCash, stormDataReduced$PROPDMG, stormDataReduced$PROPDMGEXP)
stormDataReduced$CROPCASH <- mapply(calcCash, stormDataReduced$CROPDMG, stormDataReduced$CROPDMGEXP)
stormDataReduced$TOTCASH <- stormDataReduced$PROPCASH + stormDataReduced$CROPCASH
```

Now we'll aggregate all the data down to YEAR and EVTYPE. We'll calculate the fatalities, injuries, property damage, crop damage, and total damage and use those values for our analysis.
```{r stormImpact}
stormImpact <- stormDataReduced %>%
    select(EVTYPE, YEAR, FATALITIES, INJURIES, PROPDMG, PROPDMGEXP, PROPCASH, CROPDMG, CROPDMGEXP, CROPCASH, TOTCASH ) %>%
    group_by(YEAR, EVTYPE) %>%
    summarise(totalFatalities = sum(FATALITIES), 
              totalInjuries = sum(INJURIES),
              totalPropCash = sum(PROPCASH),
              totalCropCash = sum(CROPCASH),
              totalCash = sum(TOTCASH))    

```

# Results

## Most Harmful Event Types to Public Health
Question) Across the United States, which types of events are most harmful with respect to population health?

```{r timeSeriesHealth, fig.height=15}
#melt the data to make it easier to plot
stormHealth <- melt(stormImpact[,c("YEAR","EVTYPE","totalFatalities","totalInjuries")], 
     id.vars=c("YEAR","EVTYPE"), variable.name = "impactType", value.name = "impactCount")

stormHealth %>%
    ggplot() +
        geom_line( aes(x = YEAR, y = impactCount, color = impactType), size=1) +
        facet_wrap(~ EVTYPE, ncol = 2) +
        theme_bw() +
        scale_colour_discrete(name="Impact Type", labels=c("Injuries","Fatalities")) +
        labs(x = "Year", y = "Health Impact (people)") +
        ggtitle("Storm Impact on Public Health By Event Type")    
```

Based on the plots the most harmful event types seem to be TORNADO, FLOOD, EXCESSIVE HEAT, FLASH FLOOD, TSTM WIND.  More research and investigation should be done on these types.  We can strenghten this view by looking at the cumulative  fatalaties and injuries for each event type across the entire timeframe 1995-2011.

```{r cumulativeStormImpact}
cumulativeStormImpact <- stormImpact %>%
    select(YEAR, EVTYPE, totalFatalities, totalInjuries, totalPropCash, totalCropCash, totalCash) %>%
    group_by(EVTYPE) %>%
    summarise(allFatalities = sum(totalFatalities), 
              allInjuries = sum(totalInjuries),
              allPropCash = sum(totalPropCash),
              allCropCash = sum(totalCropCash),
              allCash = sum(totalCash)
              )
```

Top 5 Event Types for Fatalities
```{r}
head(arrange(cumulativeStormImpact, desc(allFatalities)), 5)
```

Top 5 Event Types for Injuries
```{r}
head(arrange(cumulativeStormImpact, desc(allInjuries)), 5)
```

## Most Harmful Event Types to the Economy
Question) Across the United States, which types of events have the greatest economic consequences?

```{r timeSeriesEconomy, fig.height=15}

#melt data for plot and reduce Dollars scale to millions
stormEconomy <- melt(stormImpact[,c("YEAR","EVTYPE","totalPropCash","totalCropCash","totalCash")], 
     id.vars=c("YEAR","EVTYPE"), variable.name = "Type", value.name = "Dollars") %>%
    mutate(Dollars = Dollars/1000000 ) 

stormEconomy %>%
    ggplot() +
        geom_line( aes(x = YEAR, y = Dollars, color = Type), size=1) +
        facet_wrap(~ EVTYPE, ncol = 2) +
        theme_bw() +
        scale_colour_discrete(name="Impacted Type", labels=c("Property","Crops","Total")) +
        labs(x = "Year", y = "Economic Impact (Millions of $)") +
        ggtitle("Storm Impact on US Economy By Event Type")    
```

Top 5 Event Types for Property Damage
```{r}
head(arrange(cumulativeStormImpact, desc(allPropCash)), 5)
```

Top 5 Event Types for Crop Damage
```{r}
head(arrange(cumulativeStormImpact, desc(allCropCash)), 5)
```

Top 5 Event Types for Total Damage
```{r}
head(arrange(cumulativeStormImpact, desc(allCash)), 5)
```


Based on the plots and the data, the most harmful event types seem to be FLOOD, TORNADO, HAIL, FLASH FLOOD, DROUGHT.  More research and investigation should be done on these types. 


Note:  This report has been published on RPubs at http://rpubs.com/ercorne/severeweather
