---
title: "Exploratory Data Analysis Notes"
author: "Erik Cornelsen"
date: "July 26, 2016"
output:
  html_document:
    fig_caption: yes
    toc: yes
    toc_depth: 4
    number_sections: yes

---

----------------------------------------------------

## 4 - Exploratory Data Analysis (Graphics)  -  *[toc](#TOC)*

* Plots help to understand data, find patterns, resolve problems, suggest strategies for modeling, and communicate with others
* Exploratory plots are for your personal understanding, not for formal sharing.  usually ad-hoc and lots of them.
* Try to always have a background/vague question before you start plotting, since it will drive what you find.
* Find examples of other peoples graphs
    * [R Graph Gallary](http://rgraphgallery.blogspot.com/)
    * [R Bloggers](http://www.r-bloggers.com/)
* There are **3 main plotting systems**: base, lattice, ggplot2

### Visualization Principles  -  *[toc](#TOC)*

1. Show Comparisons. When giving a summary of data, always ask "compared to what?".
	1. air cleaner, compared to no air cleaner
1. Show causality, mechanism, explanation, systematic structure	
	1. what do you think is happening?
	1. symptoms improved with air cleaner? > we think it's removing particulate matter
	1. don't confuse correlation with causation, correlation simply suggests something
1. Show multivariate data (more than 2 vars)	
	1. real world has lots of things interacting, need to tell a richer story.
	1. single graphs could usually acomodate 3 variables, then make sub graphs to split on a 4th
1. Integrate multiple modes of evidence 
	1. use different ways to display the evidence (plot, table, etc)
	1. completly integrate words, numbrs,images, diagrams
	1. don't let the tool drive the analysis, you pick the right tool for the need.
1. Describe and Document Evidence in your graphics
	1. give credibility by using appropriate lables, scales, sources, etc
1. Content is King
	1. need high quality, relevance, and integrity of content (data + display)
	1. need an interesting story to tell or graphics can't help

(Source: 'Beautiful Evidence' www.edwardtufte.com)		

### Graphics Devices  -  *[toc](#TOC)*

1. a graphics device is anywhere you can make a graph appear (screen, pdf, jpg, etc)
    1. usually use screen to explore and send to other devices for final output 
    1. output of devices are usually vector or bitmap formats
        1. vector formats - resize easily.  good with lines.  not good for tons of data points.
        1. bitmap formats - don't resize. not good with lines. good with tons of data points. 
        1.  Vector Formats      |   Bitmap Format
            --------------------|-------------------------------
            pdf()               |   png() - good 4 lines and points
            svg() - good 4 web  |   jpg() - good 4 color gradients
            win.metafile()      |   tiff()
            postscript()        |   bmp()
1. Writing to a device
    1.  > pdf(file='name')  
        > ...code to create plot here...  
        > dev.off()  # close device  
1. device functions
    1. dev.cur() - which device is currently active
    1. dev.set() - choose an existing device to use
    1. dev.copy()
    1. dev.compy2pdf()


### Base Plotting System  -  *[toc](#TOC)*

* "Artists Palate" approach where you create each piece from blank canvas
* Good for fine grain control as you have to specify everything. useful when creating publication level plots
* has a 2 phase paradigm. 1) initialize/generate plot  2) annotate the existing plot

1. core functions
    1. hist() - shows distribution of a single variable
    1. plot() - basic x,y plot
    1. boxplot() - boxes at different levels
1. other base functions
    1. plot()
    1. lines() - connect the dots
    1. points() - add points to plot
    1. text() - add text inside plot
    1. title() - add label outside plot
    1. mtext() - margin text
    1. axis() - axis labels & ticks
    1. abline() - add a 'level' line to the plot
1. key common paramaters
    1. type - one character type of plot (line, bar, smoothline, points)
    1. pch - plotting symbol
    1. lty - line type
    1. lwd - line width
    1. col - color (numeric, string, hex, colors() )
    1. xlab - xlabel
    1. ylab - y label
    1. main,main2 - main title/label of the entire plot
1. Global paramaters using **par()**
    1. **TIP:** save default par values to variable before you begin changes, that way you can restore them if needed.
    1. las - axis label orientation
    1. bg - background color
    1. mar - margin size
    1. oma - outer margin size
    1. mfrow - plots per row
    1. mfcol - plots per column
    

### Lattice Plotting System  -  *[toc](#TOC)*

* is good for creating lots of plots at once on high dimensional data
* large number of plots constructed with single function call but annotation can be tricky

1. Functions
    1. xyplot(y ~ x | f * g, dataframe)
        1. uses formula notation (reason for '~')
        1. reads as: "look @ scatterplot of y & x for every level of f & g"
        1. returns an objecyt of class 'trellis' that must be printed.  R's autoprinting takes care of it.
    1. Panel functions are used to customize inside each panel
        1. panel.xyplot() - default needed to make data appear
        1. panel.abline() - overlay specified line
        1. panel.lmline() - overlay linear regression line (best fit)

  
### ggplot2 Plotting System  -  *[toc](#TOC)*

* implements 'Grammer of Graphics' concept 
* shorten the distance from the mind to the page

1. visiualizations are composed of different geomitires that work with each other. 
1. Some of the basic components: data > geometries/layers > aesthetics
    + data
	+ geometries / layers (points, lines, bars)
	+ aesthetics (x, y, size, color, shape, alpha, etc)
	+ scales (numeric, color, size)
1. important to use factor variables to slice and dice subsets of the data (label them well)
1. facets are like panels in lattice, separate displays of different points
1. **qplot()** - like base plot() function, built on top of ggplot2()
    ```{r qplot}
    library(ggplot2)
    qplot(x = displ, y = hwy, data = mpg, color = drv)
    qplot(displ,hwy,data=mpg,geom=c("point","smooth"))
    qplot(hwy,data=mpg,fill=drv)
    #ggsave("plot.png")
    ```
1. [**ggplot()**](docs.ggplot2.org) - core function
    1. **Prefered**: is very flexible. more raw > undersand what you're doing more than using qplot()
    1. SAMPLE CODE for ggplot2
    
    `
        library(ggplot2)
        
        ggplot(data=diamonds) +
          geom_point(aes(x=carat, y=price, color=clarity)) +
          geom_smooth(aes(x=carat, y=price)) +
          facet_grid(color ~ cut)
        
        ggplot(data=diamonds) +
          geom_point(aes(x=carat, y=depth, size=price))
        
        ggplot(diamonds) +
          geom_point(aes(x=carat, y=table, color=price, shape=clarity, alpha=depth))
        
        #-- Advanced Thought
        #ggplot expects your dataframe to be 'tidy':
        # - each row is an observation
        # - each column is an attribute/variable/dimension
        
        #But, you can "melt" the dataframe into a longer shape:
        #	diamonds$id <- seq(1, nrow(diamonds))
        #	d2 <- melt(data=diamonds, id.vars=c("id"), measure.vars=c("carat", "depth", "table", "price"))
    `

### pairs Plots  -  *[toc](#TOC)*

1. pairs(DT)
1. GGally has a more detailed pairs plot
    1. any colour variable has to be a factor
    1. you can specify columns, behavior of diagonal entries, and labels
    1. `?ggpairs` has good examples at the bottom that are very helpful for visualizing some of the possibilities.    
```{r}
library(GGally)
data(iris)
ggpairs(iris, colour='Species', alpha=0.4)
data(tips, package = "reshape")
ggpairs(tips[,1:3])
ggpairs(
  tips[,c(1,3,4,2)],
  upper = list(continuous = "density", combo = "box"),
  lower = list(continuous = "points", combo = "dot")
)
```
1. use ggplot2 normally
```
 ggplot(data, aes(x=id, y=value)) + geom_point() + facet_grid(.~variable). 
```
